---
title: "Practical Machine Learning Project"
author: "Scott Posey"
date: "Saturday, May 21, 2016"
output: html_document
---

## Project Assignment Background:

"Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement ??? a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset)."


## Data Sources:

The training and test data for this project are available here:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

The original source of the data is: http://groupware.les.inf.puc-rio.br/har. If you use the document you create for this class for any purpose please cite them as they have been very generous in allowing their data to be used for this kind of assignment.


## Objective:

To predict the manner in which the subjects performed the exercise. This is the "classe" variable in the training set.

## Libraries Used: 

```{r ,message=FALSE}
library(abind)
library(arm)
library(caret)
library(kernlab)
library(klaR)
library(rattle)
library(randomForest)
library(rpart)
```

## Reproducibility:

Random Seed 

```{r}
set.seed(21165)
```

## Import Data:

Setting train and testing datasets' urls:

```{r}
urlTrain <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
urlTest <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
```

## File retrieval and reading:

```{r}
csvTrain <- "pml-training.csv"

if (file.exists(csvTrain)) {
        train <- read.csv(csvTrain, na.strings=c("NA","#DIV/0!",""))
} else { 
        download.file(urlTrain,csvTrain)
        train <- read.csv(csvTrain, na.strings=c("NA","#DIV/0!",""))
        }                           

csvTest <-  "pml-testing.csv"

if (file.exists(csvTest)) {
        test <- read.csv(csvTest, na.strings=c("NA","#DIV/0!",""))
} else { 
        download.file(urlTest,csvTest)
        test <- read.csv(csvTest, na.strings=c("NA","#DIV/0!",""))
}   
```

## Clean Training Set Data:

Coherence Check

```{r}
all.equal(colnames(test)[1:length(colnames(test))-1], colnames(train)[1:length(colnames(train))-1])
```
```{r ,results='hide'}
summary(train)
summary(test)
```
Remove variables with low variance

```{r ,results='hide'}
nearzero <- nearZeroVar(train, saveMetrics = TRUE)
train <- train[, !nearzero$nzv]
```
Remove variables with more than 50% missing values (NA)

```{r ,results='hide'}
toberem <- sapply(colnames(train), function(x) if(sum(is.na(train[, x])) > 0.50*nrow(train))    {return(TRUE)
}else{
return(FALSE)
}
)
train <- train[, !toberem]
```

Remove data acquisition variables (eg id, timestamps, names, etc.) 
```{r ,results='hide'}
train <- train[, -(1:6)]
```

## Correlation analysis:

```{r ,results='hide'}
Hcorr <- caret::findCorrelation(cor(train[, -53]), cutoff=0.8)
names(train)[Hcorr]
```

Post cleaning variable set
```{r }
names(train)
```
## Model Specification & Cross Validation

```{r }
tc <- trainControl(method = "cv", number = 7, verboseIter=FALSE , preProcOptions="pca", allowParallel=TRUE)
```

Six models are estimated:

* Random forest
* Logit Boosted Model.
* Bayes Generalized Linear Model
* Support Vector Machine - Radial
* Support Vector Machine - Linear
* Neural Net

```{r }
rf <- train(classe ~ ., data = train, method = "rf", trControl= tc)
logitboost <- train(classe ~ ., data = train, method = "LogitBoost", trControl= tc)
bayesglm <- train(classe ~ ., data = train, method = "bayesglm", trControl= tc)
svmr <- train(classe ~ ., data = train, method = "svmRadial", trControl= tc)
svml <- train(classe ~ ., data = train, method = "svmLinear", trControl= tc)
NN <- train(classe ~ ., data = train, method = "nnet", trControl= tc, verbose=FALSE)
```

### Accuracy comparision

```{r }
model <- c("Random Forest","LogitBoost", "Bayes GLM", "SVM (radial)","SVM (linear)","Neural Net")
Accuracy <- c(max(rf$results$Accuracy),
              max(logitboost$results$Accuracy),
              max(bayesglm$results$Accuracy),
              max(svmr$results$Accuracy),
              max(svml$results$Accuracy),
              max(NN$results$Accuracy)
             )
        
Kappa <- c(max(rf$results$Kappa),
           max(logitboost$results$Kappa),
           max(bayesglm$results$Kappa),
           max(svmr$results$Kappa),
           max(svml$results$Kappa),
           max(NN$results$Kappa),
          )  

performance <- cbind(model,Accuracy,Kappa)
knitr::kable(performance)
```

Random forest and SVM(radial) provide the best results and will provide the predictions for the submission. Even if the Out of sample error cannot be estimated exactly, the in-sample error obtained through cross-validation is calculated over different test sets and should provide a better estimate of out-of sample error with respect to the case of no cross-validation.

Prediction of "classe" variable for the test set
```{r}
rfPred <- predict(rf, test)
svmrPred <- predict(svmr, test)
```

Checking if the models give same predictions
```{r}
prediction <- data.frame(cbind(rfPred, svmrPred))
prediction$same <- with(prediction, rfPred == svmrPred)
colnames(prediction) <- c("Random Forest", "SVM (radial)", "Same Prediction")
knitr::kable(prediction)
```

Generation of the files to be submitted is made through the provided function
```{r}
pml_write_files = function(x){
        n = length(x)
        for(i in 1:n){
                filename = paste0("problem_id_",i,".txt")
                write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
        }
}

pml_write_files(rfPred)
pml_write_files(svmrPred)
```

## Conclusion:

The random forest model provides an outstanding accuracy and, accordingly, the predictions for the test set were correct in 100% of the cases.
